---
title: "Generalized linear models - VT26 "
subtitle: "Inlämningsuppgift 1, 2.5 hp"
author: 
- David Müller
- Filip Ström
- Joel Söderman
date: last-modified
format: 
  pdf: default
  html:
    self-contained: true
language: 
  title-block-author-single: " "
crossref-fig-title: "Figur"
title-block-banner-color: Primary
title-block-published: "Publicerad"
author-title: "Gruppmedlemmar"
title-block-banner: true
editor: source
---

```{r}
#| output: false
#| echo: false
library('foreign')
library('haven') 
library('labelled')

load(url("https://raw.githubusercontent.com/johankoskinen/ST304G/main/data/GSS2022selection.RData"))

load(url("https://raw.githubusercontent.com/johankoskinen/ST304G/main/data/GSS2024selection.RData"))
```

# Introduction
This report uses data from the General Social Survey (GSS) to study two different types of outcomes with generalized linear models. In Part A, we examine which factors are associated with voting for Donald Trump in the 2020 presidential election. In Part B, we analyse the reported number of sexual partners and discuss suitable modelling strategies for a partly censored and highly skewed response variable.

The purpose of the assignment is to apply generalized linear models to real survey data, compare different model specifications and link functions, and interpret the results in substantive terms. A central aim is not only to estimate statistical models, but also to evaluate how well they fit the data and whether the conclusions are robust to different modelling choices.

The report is therefore structured around model estimation, interpretation, and comparison. In Part A, we focus on binary response models for Trump voting and examine whether the role of sex changes across survey years. In Part B, we first consider a Poisson model for count data and then explore alternative binary models in order to better capture the structure of the outcome.

### Note on AI use 
We used AI in a limited way to help format the output in the report, mainly by suggesting how to use knitr::kable() and similar commands to present summary() and anova() results more clearly. 

# Task 1
In this task, we examine whether sex, gun ownership, and education are associated with voting for Donald Trump in the 2020 presidential election. We estimate the same logistic regression model using both the 2022 and the 2024 GSS data. This allows us to compare the results across the two surveys and to test whether the association between being female and voting for Trump changed over time.

```{r}
#| echo: false
model_22 <- glm((pres20 == 2) ~ 
                  I(sex == 2) + 
                  I(owngun == 1) + 
                  I(degree %in% c(4,5)),
               family = "binomial",
               data = GSSdata22)

knitr::kable(summary(model_22)$coefficients, digits = 3, caption = "Model 2022")
```
For the 2022 model, gun ownership is positively associated with voting for Trump, while having a Bachelor’s or Graduate degree is negatively associated with the outcome. The coefficient for being female is also negative, and in this model the effect is statistically significant.

```{r}
#| echo: false
#| 
model_24 <- glm(pres20 == 2 ~ 
                   I(sex == 2) + 
                   I(owngun == 1) +
                   I(degree %in% c(4,5)), 
                 data = GSSdata24,
                 family = "binomial")


knitr::kable(summary(model_24)$coefficients, digits = 3, caption = "Model 2024")

```
The 2024 model shows a similar overall pattern. Gun ownership is again positively associated with voting for Trump, while higher education is negatively associated with the outcome. The coefficient for being female remains negative, but here it is not statistically significant.


### Pooled Model with Interaction
To formally test whether the association between being female and voting for Trump changed over time, we then created a pooled dataset that combines the observations from 2022 and 2024. In the code, we first selected the variables needed for the analysis, then added a new variable called year to identify which survey year each observation comes from, and finally stacked the two datasets together using rbind(). In the combined model, we included an interaction term between I(sex == 2) and I(year == 2024). This interaction term is the key part of the code, because it tests whether the effect of being female differs between 2022 and 2024, rather than simply estimating one common effect across both years.
```{r}
#| echo: false
vars_needed <- c("pres20", "sex", "owngun", "degree")

GSSdata22_sub <- GSSdata22[, vars_needed]
GSSdata24_sub <- GSSdata24[, vars_needed]

GSSdata22_sub$year2024 <- 0
GSSdata24_sub$year2024 <- 1

GSSdata <- rbind(GSSdata22_sub, GSSdata24_sub)

model <- glm(
  I(pres20 == 2) ~
    I(sex == 2) +
    I(owngun == 1) +
    I(degree %in% c(4, 5)) +
    year2024 +
    I(sex == 2):year2024,
  data = GSSdata,
  family = binomial
)

knitr::kable(summary(model)$coefficients, digits = 3, caption = "Pooled Model")

```
The p-value for the interaction term is 0.787. Since this p-value is large, there is no statistical evidence that the association between being female and voting for Trump changed between the two years. Therefore, we cannot reject the null hypothesis that the effect of being female is the same in 2022 and 2024. 

# Task 2:
We estimated four models: a small and a large specification, each under both a logit and a probit link. To make the model comparisons fair, all four models were estimated on the same complete cases. The smaller models include only sex, gun ownership, and education, while the larger models also include region, race, type of community, and logged income. We compare the models by examining the estimated coefficients, using ANOVA to compare the smaller and larger specifications, and using AIC to compare overall model fit. We also verify that all four models are estimated on the same number of observations.


```{r}
#| output: false
#| echo: false
vars_needed <- c(
  "pres20", "sex", "owngun", "degree",
  "region", "race", "res16", "realinc"
)

GSS_clean <- na.omit(GSSdata24[, vars_needed])

m_small_logit <- glm(
  pres20 == 2 ~
    I(sex == 2) +
    I(owngun == 1) +
    I(degree %in% c(4, 5)),
  data = GSS_clean,
  family = binomial(link = "logit")
)

m_big_logit <- glm(
  pres20 == 2 ~
    I(sex == 2) +
    I(owngun == 1) +
    I(degree %in% c(4, 5)) +
    factor(region) +
    factor(race) +
    factor(res16) +
    log(realinc),
  data = GSS_clean,
  family = binomial(link = "logit")
)

m_small_probit <- glm(
  pres20 == 2 ~
    I(sex == 2) +
    I(owngun == 1) +
    I(degree %in% c(4, 5)),
  data = GSS_clean,
  family = binomial(link = "probit")
)

m_big_probit <- glm(
  pres20 == 2 ~
    I(sex == 2) +
    I(owngun == 1) +
    I(degree %in% c(4, 5)) +
    factor(region) +
    factor(race) +
    factor(res16) +
    log(realinc),
  data = GSS_clean,
  family = binomial(link = "probit")
)

```

### Small logit model

```{r}
#| echo: false
knitr::kable(summary(m_small_logit)$coefficients, 
             caption = "Small logit model")
```
The small logit model shows that gun ownership is positively associated with voting for Trump, while having a Bachelor’s or Graduate degree is negatively associated with the outcome. The coefficient for being female is negative, but it is not statistically significant, which means that we do not find clear evidence of a gender effect in the smaller logit specification.

### Big logit model
```{r}
#| echo: false
knitr::kable(summary(m_big_logit)$coefficients, 
             caption = "Big logit model")

```
In the larger logit model, the main pattern remains similar. Gun ownership is still positively associated with voting for Trump, while higher education is negatively associated with the outcome. After adding region, race, type of community, and income, the model captures more variation in the data, suggesting that these background variables are also important for explaining Trump voting.


```{r}
#| echo: false
knitr::kable(summary(m_small_probit)$coefficients, 
             caption = "Small probit model")
```
The small probit model gives results that are very similar to the small logit model. Gun ownership is again positively associated with voting for Trump, while higher education is negatively associated with the outcome. The effect of being female remains negative, but not statistically significant.

```{r}
#| echo: false
knitr::kable(summary(m_big_probit)$coefficients, 
             caption = "Big probit model")
```
The large probit model also shows a pattern similar to the large logit model. The signs and significance levels of the main variables are broadly consistent across the two link functions, which suggests that our conclusions are not very sensitive to whether we use a logit or a probit link.

```{r}
#| echo: false
knitr::kable(anova(m_small_logit, m_big_logit), 
             caption = "ANOVA: Small vs Large logit Model") 
      
```


```{r}
#| echo: false
knitr::kable(anova(m_small_probit, m_big_probit), 
             caption = "ANOVA: Small vs Large probit Model")
```
The ANOVA results show that the larger model improves the fit significantly compared with the smaller model, both for the logit and probit specifications. This suggests that the additional variables contribute useful explanatory information and should be included in the final model.

To compare the overall fit of the four models, we use the Akaike Information Criterion (AIC):
$$
AIC = 2p - 2 \log(L(\hat{\beta}))
$$
where $p$ is the number of parameters in the model and $(L(\hat{\beta})$ is the likelihood evaluated at the estimated parameter values.Lower AIC values indicate a better balance between model fit and model complexity.

```{r}
#| echo: false
aic_tab <- AIC(m_small_logit, m_big_logit, m_small_probit, m_big_probit)
rownames(aic_tab) <- c("Small logit", "Large logit", "Small probit", 
                       "Large probit")

knitr::kable(aic_tab[, c("df", "AIC")], caption = "AIC")


```
Overall, the results suggest that demographic and social background variables are important for explaining Trump voting in 2020. In particular, gun ownership is associated with a higher probability of voting for Trump, while higher education is associated with a lower probability. Both the ANOVA results and the AIC values indicate that the larger models fit the data better than the smaller models. Since the large logit model has the lowest AIC among the four models and is easier to interpret than the corresponding probit model, we choose it as our preferred final model. The similarity between the logit and probit results also suggests that our main conclusions are robust to the choice of link function.

```{r}
#| output: false
#| echo: false
nobs(m_small_logit)
nobs(m_big_logit)
nobs(m_small_probit)
nobs(m_big_probit)
```

# Task 3 
To answer this task, we first created a new dataset where we only kept observations with complete information on the variables used in our models: sex, gun ownership, education, region, race, type of community, and income. We did this by using complete.cases(...), which checks which rows have no missing values (NA) on these variables. This step is necessary because the models cannot make predictions for individuals if some of the explanatory variables are missing.

After this, we selected the individuals in this cleaned dataset who belong to the Midwest and who did not answer the question about the 2020 presidential election. In the dataset, Midwest is coded as region == 2, so we use this condition to identify that region. At the same time, we use is.na(pres20) to find the non-respondents, that is, the individuals with a missing value on the outcome variable. The symbol & is used to combine these conditions, meaning that both must be true at the same time: the person must both live in the Midwest and have a missing value on pres20.

Finally, we used our two models, the larger logit model and the larger probit model, to predict the probability that each individual in this group voted for Trump. We then took the average of these predicted probabilities. This gives us a model-based estimate of the proportion of Trump voters among the non-respondents in the Midwest.


```{r}
#| echo: false
GSS_pred <- GSSdata24[
  complete.cases(GSSdata24[, c("sex", "owngun", "degree", "region", 
                               "race", "res16", "realinc")]),
]

midwest_nr <- GSS_pred[GSS_pred$region == 2 & is.na(GSS_pred$pres20), ]

logit_est <- mean(predict(m_big_logit, newdata = midwest_nr, 
                          type = "response"))
probit_est <- mean(predict(m_big_probit, newdata = midwest_nr, 
                           type = "response"))

pred_tab <- data.frame(
  Model = c("Large logit", "Large probit"),
  Estimated_percent = c(logit_est, probit_est) * 100
)

knitr::kable(pred_tab, digits = 1, caption = "predictions for non-respondents in the Midwest")


```
The large logit model estimates that approximately 31.6 percent of the non-respondents in the Midwest voted for Trump, while the large probit model gives a very similar estimate of 31.8 percent. Since the two estimates are almost identical, the result appears robust to the choice of link function. We therefore conclude that the predicted proportion of Trump voters among non-respondents in the Midwest is about 32 percent.


# Part B
### Task 4
The variable "partners" refers to the number of sexual partners reported by the respondent in the survey. The outcome of the variable is partially censored, some observed values are recorded as interval-censored categories. For example, the value "5" corresponds to the interval 5-10 sexual partners.
```{r}
table(GSSdata24$partners , useNA = 'always')
```
The issue of partially censored outcomes is handled in different ways. The value “5”, which represents the interval 5–10, is set to 7 as it corresponds approximately to the midpoint of the interval. The value “6”, which represents the open-ended interval 11+, is set to 12 as a conservative approximation. The value “9” is treated as NA, as no reasonable numerical approximation can be made without introducing strong arbitrary assumptions.

Sedermera valde vi att ta bort de saknade värderna från vårt dataset då GLM modeller kräver observerade responsvärden. Dessutom så baseras GLM estimeringar på likelihoodfunktioner vilka kräver observerade utfall. GLM-modeller kan heller inte modellera censorerad data vilket således gör att vi måste approximera intervallen för att möjliggöra likelihood estimationer. 

Mätfel:
I och med att vi approximerar de grupperade variablerna så tillkommer det mätfel, dessa kan vi inte udgås men för att möjliggöra GLM krävs dessa antaganden. Detta då vi dels vidhåller samma ordning i datan samt att vi förhåller oss till "top coded" problematiken. Hämtat från Wikipedia (2020-12-02, "Top-coded"): "To help overcome this problem, CPS provides the mean value of top coded values". 

Vi skapar även en binär variabel "married" för att kunna undersöka om äktenskap påverkar antal partners: TRUE = gift, FALSE = ej gift. 

```{r}
dat_24 <- GSSdata24
dat_24$partners_num <- dat_24$partners

dat_24$partners_num[dat_24$partners == 5] <- 7
dat_24$partners_num[dat_24$partners == 6] <- 12
dat_24$partners_num[dat_24$partners == 9] <- NA

# Kanske lite längre förklaring till denna kod-del. 
dat_24 <- subset(dat_24, !is.na(partners_num))

dat_24$married <- dat_24$marital == 1
```

### Länkfunktion 1 - Poission Log-link:

Eftersom att responsvariabeln representerar antal har vi valt att använda en poisson regression med log länk. 

$$log(\mu) = \beta_0 + \beta_1 * marriedTRUE$$
där:
$$E(Y) = \mu $$
Alltså, vi modellerar förväntat antal partners. 
```{r}
pois_1 <- glm(partners_num ~ married,
              family = poisson(link = "log"),
              data = dat_24
)

summary(pois_1)

b <- coef(pois_1)

# Förväntat antal partners för en ogift person
mu_unmarried <- exp(b[1]) # = 1.124277

# Förväntat antal partners vid gift
mu_married <- exp(b[1] + b[2]) # = 1.037569

list("partners vid ogift" = mu_unmarried, "partners vid gift" = mu_married)
```

Tolkning av koefficienter:

Intercept (b[1] = 0.11714): Detta är det logaritmerade "mu" för en ogift individ När vi transformerar detta (exp) får vi ut att det förväntade antalet partners för en ogift individ uppgick till 1.124277 stycken. 

marriedTRUE (-0.08026): Detta anger skillnaden mellan att vara ogift kontra gift. Vid transformation av detta värdet tillsammans med interceptet får vi ut ett förväntat värde på 1.0376 Sätt ur en procentuell vinkel så kan vi säga att gifta personer i genomsnitt har $8 \%$ färre förväntade partners än ogifta. (Se nedan varför vi kan göra denna tolkning)

p-värde (0.0426): marriedTRUE är signifikant på en 95% nivå, således kan vi med 95% säkerhet konstatera att gifta individer har färre partners.

Modellantagande Poisson:

$$ Var(Y) = E(Y)$$
```{r}
mean(dat_24$partners_num)
var(dat_24$partners_num)
```
Variansen (2.7833) överstiger medelvärdet (1.0899) vilket indikerar att poissonantagandet om lika varians och medelvärde är begränsat, detta kallas för overdispersion. Konsekvensen av detta är att standardfelet för våra koefficienter högst troligt är underskattatade. Här hade det möjligtvis varit mer lämpligt med en negativ binomial modell som tillåter att detta antagande inte hålls. 

### Länkfunktion 2 - Logit-modeller:

$$Odds = e^{\beta}$$
$$Sannolikhet = \frac{Odds}{1 + Odds}$$
Vi skapar 2 stycken binära variabler:

- Minst en partner (any_partner)

- Fler än en partner (many_partners)
```{r}
dat_24$any_partner <- dat_24$partners_num > 0

dat_24$many_partners <- dat_24$partners_num > 1
```

```{r}
# Logit modell - minst en partner:
logit_any <- glm(any_partner ~ married, 
                 family = binomial(link = "logit"), 
                 data = dat_24
                 )
knitr::kable(summary(logit_any)$coefficients)

b_2 <- coef(logit_any)

odds_any_unmarried <- exp(b_2[1]) # 1.46
odds_any_married <- exp(b_2[1] + b_2[2]) # 7.238

prob_any_unmarried <- odds_any_unmarried / (1 + odds_any_unmarried) # 0.5934
prob_any_married <- odds_any_married / (1 + odds_any_married) # 0.8786

list("Prob if unmarried" = prob_any_unmarried, 
     "Prob if married" = prob_any_married)
```
Tolkning av koefficienter:

Intercept (0.378): Är vårt logodds för ogifta individer som har haft minst en partner. Vid transformation av detta får vi ut oddset som uppgick till 1.46. Sedermera omvandlar vi detta till en sannolikhet som säger oss att ogifta personer har ca $59 \%$ sannolikhet att ha haft minst en partner

marriedTRUE: Är förändringen i logodds mellan individer som är ogifta och gifta. Vi transformerar detta till odds genom att addera $\beta_1$ (1.6) med interceptet för att sedan exponentiera det (Odds = 7.238). Sedan omvandlar vi oddset till en sannolikhet (0.8786). Här kan det konstateras att gifta individer har ca $88\%$ sannolikhet att ha haft minst en partner, vilket således är mycket högre än de ogifta individerna. Notering: I och med att koeficienten är positiv vet vi att "vara gift" ökar sannolikheten att minst haft en partner, värt att nämna är att detta resultat är ytterst signifikant (p-värde: 2e-16). 

```{r}
# Logit modell - fler än en partner:
logit_many <- glm(many_partners ~ married, 
                 family = binomial(link = "logit"), 
                 data = dat_24
                 )
knitr::kable(summary(logit_many)$coefficients)

b_3 <- coef(logit_many)

odds_many_unmarried <- exp(b_3[1]) # 0.19076
odds_many_married <- exp(b_3[1] + b_3[2]) # 0.039

prob_many_unmarried <- odds_many_unmarried / (1 + odds_many_unmarried) # 0.16
prob_many_married <- odds_many_married / (1 + odds_many_married) # 0.0376

list("Prob if unmarried" = prob_many_unmarried, 
     "Prob if married" = prob_many_married)
```
Tolkning av koefficienter:

Intercept (-1.65673): Är vårt logodds för ogifta individer som har haft fler än en partner. Vid transformation av detta får vi ut oddset som uppgick till 0.19076. Sedermera omvandlar vi detta till en sannolikhet som säger oss att ogifta personer har ca $16 \%$ sannolikhet att ha haft fler än en partner.

marriedTRUE: Är förändringen i logodds mellan individer som är ogifta och gifta. Vi transformerar detta till odds (0.039) och sedan omvandlar detta till en sannolikhet (0.0376). Här kan det konstateras att gifta individer har ca $3.76\%$ sannolikhet att ha haft fler än en partner, vilket således är mycket lägre än de ogifta individerna. Notering: I och med att koeficienten är negativ vet vi att "vara gift" minskar sannolikheten att ha haft fler än en partner, värt att nämna är att detta resultat är ytterst signifikant (p-värde: 2e-16 ≈ 0). 

### Länkfunktion 3 - Probit, samt jämförelse med logit
```{r}
#| echo: false
probit_any <- glm(any_partner ~ married, 
                  family = binomial(link = "probit"),
                  data = dat_24
                  )

probit_many <- glm(many_partners ~ married, 
                  family = binomial(link = "probit"),
                  data = dat_24
                  )
knitr::kable(summary(probit_any)$coefficients,
             caption = "Probit any")
knitr::kable(summary(logit_any)$coefficients,
             caption = "Logit any")

knitr::kable(summary(probit_many)$coefficients,
             caption = "Probit many")
knitr::kable(summary(logit_many)$coefficients,
             caption = "Logit many")

```

Resultat:

Resultatet från probit modellerna bekräftar de tidigare logit analyserna och påvisar att slutsatserna är rimliga med avseende på valet av länkfunktion för de binära responserna (logit, probit). Både logit och probit ger samma tecken (+ respektive - för coeffisienterna) och lika signifikansnivåer. Äktenskap ökar sannolikheten att ha minst en partner, men minskar sannolikheten att ha fler än en partner. Med detta kan vi konstatera att det finns ett mönster i datan och att dessa "två stegs" modeller inte är beroende av en specifik matematisk formel. 

### Prediktion (Hypotetisk respondent):
Vi betraktar en hypotetisk respondent som antingen är gift eller ogift och undersöker hur modellerna predicierar antal partners (poisson) respektive sannolikheten (logit, probit). Detta gör vi genom att skapa hypotetiska individer (new_person) som vi sedan applicerar på våra modeller i en dataframe. Detta skapar en tabell som ger oss de önskade värderna.  

Kommandot type = "response" transformerar modellens linjära prediktor tillbaka till dess originalskala. För poissonmodellen innebär det "förväntat antal partners", medan för logitmodellen innebär det sannolikhet. 

```{r}
new_person <- data.frame(married = c(FALSE, TRUE))

data.frame(married = new_person$married, 
# Poisson (antal)
        pois_count = predict(pois_1, 
        newdata = new_person, 
        type = "response"),
# Logit (any)
        logit_any = predict(logit_any,
        newdata = new_person,
        type = "response"), 
# Logit (many)
        logit_many = predict(logit_many,
        newdata = new_person,
        type = "response"), 
# Probit (any)
        probit_any = predict(probit_any,
        newdata = new_person,
        type = "response"), 
# Probit (many)
        probit_many = predict(probit_many,
        newdata = new_person,
        type = "response")
)

```
Enligt poisson modellen förväntas en ogift individ (1.1243) rapportera något fler partners kontra en gift individ (1.0376). Logit och probit modellerna predicerar att gifta individer (0.8786) har högre sannolikhet att ha haft minst en partner jämfört med ogifta individer (0.5934). Därtill kan det konstateras att gifta individer (0.0376) har lägre sannolikhet att ha haft fler än en partner kontra en ogift individ (0.16). Data framen illustrerar fem olika aspekter utifrån den binära variablen married.

###AIC, assesment of fit mellan modellerna:

För att bedöma modellernas anpassning (model of fit) använder vi Akaike Information Criterion (även kallat AIC). AIC bygger på modellens maximala likelihood och straffar sedan för komplexiteten i modellen. p är antalet parametrar och L är likelihoodfunktionen. Lägre AIC värden indikerar på på en bättre balans mellan modellanpassning och modellens komplexitet. 

$$ AIC = 2p-2 log (L(\widehat{\beta}))$$
```{r}
aic_tab_2 <- AIC(logit_any, logit_many, probit_any, probit_many)

knitr::kable(aic_tab_2, caption = "AIC Tabell")
```
Tolkning AIC:
För det första är det viktigt att belysa att modellerna inte använder samma responsvariabel. Poissonmodellen modellerar antal partners och logit/probit modellerna modellererar binära händelser såsom sannolikheten att ha haft en respektive fler än en partner. AIC värderna bör således inte tolkas som en strikt diskret jämförelse mellan samtliga modeller, utan snarare som ett mått på de enskilda modellanpassningarna. 

De låga AIC värderna för logit och probit modellerna påvisar att variationen i partnerskap bättre förstås som diskreta steg snarare än som en enskild poissonprocess. Resultatet kan således vara till grund för en tolkning där partnerskap modellereras i två steg:

i) Sannolikheten att ha haft minst en partner, och

ii) givet minst en partner, vad är sannolikheten att ha fler än en partner. 

Sammantaget visar AIC resultaten att äktenskap främst avser sannolikheten att ha en partner och minskar sannolikheten att ha fler en än partner, snarare än att avse det exakta antalet partners man har. 

### Deviance poisson, test med goodness of fit:

För att undersöka poissonmodellens anpassning genomförs ett goodness of fit test baserat på modellens deviance och residualdevians. 
```{r}
p_god_test <- pchisq(deviance(pois_1), 
                     df.residual(pois_1),
                     lower = FALSE)

p_god_test
```
Hypoteser:
$$ H_0 : Var(Y) = E(Y)$$
$$H_a : Var(Y) ≠ E(Y)$$
Nollhypotesen i detta test är att poissonmodellen är korrekt specifierad, vilket således innebär att variansen är samma som förväntade värdet. 

Det p-värde vi får ligger väldigt nära 0 vilket innebär att poissonmodellen har en betydande del overdispersion i datan, vi förkastar alltså nollhypotesen. Detta tyder på att variationen i datan är större än vad en enskild poissonmodell kan förklara. Detta motiverar ytterliggare en binär flerstegs modell och att poissonmodellen bör undvikas.  

### Barplot över antal partners:
```{r}
#| fig-width: 5
#| fig-height: 3
#| out-width: "50%"
#| fig-align: center
#| echo: false

barplot(table(dat_24$partners_num), 
     main = "Fördelning över # partners_num", 
     xlab = "Antal partners", 
     ylab = "frekvens", 
     col = "lightblue"
     )
```

Fördelningen av antal partners är högerskev, där många tillfrågade rapporterat låga värden och få höga värden. Detta innebär att att sambandet mellan äktenskap och antalet partners inte är homogent över hela fördelningen. Denna analys blir däremot lite tvetydig då viss del av datan är censurerad. Vi antar däremot en högerskev fördelning. 

Med detta i åtanke samt AIC utvärderingen och goodness of fit testet så vidhåller vi den mer effektiva uppdelade modellen i) och ii). 

Skillnaden mellan att ha minst en partner och fler än en visar således att äktenskap främst är relaterad till sannolikheten att överhuvudtaget ha en partner, snarare än till ett högre antal partners bland individer som redan haft en partner. I och med detta kan en uppdelad modellstruktur motiveras ytterliggare. 

### Zelterman:

Detta kan kopplas till Zeltermans modell, som fokuserar på:





