---
title: "Generalized linear models - VT26 "
subtitle: "Inlämningsuppgift 1, 2.5 hp"
author: 
- David Müller
- Filip Ström
- Joel Söderman
date: last-modified
format: 
  pdf: default
  html:
    self-contained: true
language: 
  title-block-author-single: " "
crossref-fig-title: "Figur"
title-block-banner-color: Primary
title-block-published: "Publicerad"
author-title: "Gruppmedlemmar"
title-block-banner: true
editor: source
---

```{r}
#| output: false
#| echo: false
library('foreign')
library('haven') 
library('labelled')

load(url("https://raw.githubusercontent.com/johankoskinen/ST304G/main/data/GSS2022selection.RData"))

load(url("https://raw.githubusercontent.com/johankoskinen/ST304G/main/data/GSS2024selection.RData"))
```

# Introduction
This report uses data from the General Social Survey (GSS) to study two different types of outcomes with generalized linear models. In Part A, we examine which factors are associated with voting for Donald Trump in the 2020 presidential election. In Part B, we analyse the reported number of sexual partners and discuss suitable modelling strategies for a partly censored and highly skewed response variable.

The purpose of the assignment is to apply generalized linear models to real survey data, compare different model specifications and link functions, and interpret the results in substantive terms. A central aim is not only to estimate statistical models, but also to evaluate how well they fit the data and whether the conclusions are robust to different modelling choices.

The report is therefore structured around model estimation, interpretation, and comparison. In Part A, we focus on binary response models for Trump voting and examine whether the role of sex changes across survey years. In Part B, we first consider a Poisson model for count data and then explore alternative binary models in order to better capture the structure of the outcome.

### Note on AI use 
We used AI in a limited way to help format the output in the report, mainly by suggesting how to use knitr::kable() and similar commands to present summary() and anova() results more clearly. 

# Task 1
In this task, we examine whether sex, gun ownership, and education are associated with voting for Donald Trump in the 2020 presidential election. We estimate the same logistic regression model using both the 2022 and the 2024 GSS data. This allows us to compare the results across the two surveys and to test whether the association between being female and voting for Trump changed over time.

```{r}
#| echo: false
model_22 <- glm((pres20 == 2) ~ 
                  I(sex == 2) + 
                  I(owngun == 1) + 
                  I(degree %in% c(4,5)),
               family = "binomial",
               data = GSSdata22)

knitr::kable(summary(model_22)$coefficients, digits = 3, caption = "Model 2022")
```
For the 2022 model, gun ownership is positively associated with voting for Trump, while having a Bachelor’s or Graduate degree is negatively associated with the outcome. The coefficient for being female is also negative, and in this model the effect is statistically significant.

```{r}
#| echo: false
#| 
model_24 <- glm(pres20 == 2 ~ 
                   I(sex == 2) + 
                   I(owngun == 1) +
                   I(degree %in% c(4,5)), 
                 data = GSSdata24,
                 family = "binomial")


knitr::kable(summary(model_24)$coefficients, digits = 3, caption = "Model 2024")

```
The 2024 model shows a similar overall pattern. Gun ownership is again positively associated with voting for Trump, while higher education is negatively associated with the outcome. The coefficient for being female remains negative, but here it is not statistically significant.


### Pooled Model with Interaction
To formally test whether the association between being female and voting for Trump changed over time, we then created a pooled dataset that combines the observations from 2022 and 2024. In the code, we first selected the variables needed for the analysis, then added a new variable called year to identify which survey year each observation comes from, and finally stacked the two datasets together using rbind(). In the combined model, we included an interaction term between I(sex == 2) and I(year == 2024). This interaction term is the key part of the code, because it tests whether the effect of being female differs between 2022 and 2024, rather than simply estimating one common effect across both years.
```{r}
#| echo: false
vars_needed <- c("pres20", "sex", "owngun", "degree")

GSSdata22_sub <- GSSdata22[, vars_needed]
GSSdata24_sub <- GSSdata24[, vars_needed]

GSSdata22_sub$year2024 <- 0
GSSdata24_sub$year2024 <- 1

GSSdata <- rbind(GSSdata22_sub, GSSdata24_sub)

model <- glm(
  I(pres20 == 2) ~
    I(sex == 2) +
    I(owngun == 1) +
    I(degree %in% c(4, 5)) +
    year2024 +
    I(sex == 2):year2024,
  data = GSSdata,
  family = binomial
)

knitr::kable(summary(model)$coefficients, digits = 3, caption = "Pooled Model")

```
The p-value for the interaction term is 0.787. Since this p-value is large, there is no statistical evidence that the association between being female and voting for Trump changed between the two years. Therefore, we cannot reject the null hypothesis that the effect of being female is the same in 2022 and 2024. 

# Task 2:
We estimated four models: a small and a large specification, each under both a logit and a probit link. To make the model comparisons fair, all four models were estimated on the same complete cases. The smaller models include only sex, gun ownership, and education, while the larger models also include region, race, type of community, and logged income. We compare the models by examining the estimated coefficients, using ANOVA to compare the smaller and larger specifications, and using AIC to compare overall model fit. We also verify that all four models are estimated on the same number of observations.


```{r}
#| output: false
#| echo: false
vars_needed <- c(
  "pres20", "sex", "owngun", "degree",
  "region", "race", "res16", "realinc"
)

GSS_clean <- na.omit(GSSdata24[, vars_needed])

m_small_logit <- glm(
  pres20 == 2 ~
    I(sex == 2) +
    I(owngun == 1) +
    I(degree %in% c(4, 5)),
  data = GSS_clean,
  family = binomial(link = "logit")
)

m_big_logit <- glm(
  pres20 == 2 ~
    I(sex == 2) +
    I(owngun == 1) +
    I(degree %in% c(4, 5)) +
    factor(region) +
    factor(race) +
    factor(res16) +
    log(realinc),
  data = GSS_clean,
  family = binomial(link = "logit")
)

m_small_probit <- glm(
  pres20 == 2 ~
    I(sex == 2) +
    I(owngun == 1) +
    I(degree %in% c(4, 5)),
  data = GSS_clean,
  family = binomial(link = "probit")
)

m_big_probit <- glm(
  pres20 == 2 ~
    I(sex == 2) +
    I(owngun == 1) +
    I(degree %in% c(4, 5)) +
    factor(region) +
    factor(race) +
    factor(res16) +
    log(realinc),
  data = GSS_clean,
  family = binomial(link = "probit")
)

```

### Small logit model

```{r}
#| echo: false
knitr::kable(summary(m_small_logit)$coefficients, 
             caption = "Small logit model")
```
The small logit model shows that gun ownership is positively associated with voting for Trump, while having a Bachelor’s or Graduate degree is negatively associated with the outcome. The coefficient for being female is negative, but it is not statistically significant, which means that we do not find clear evidence of a gender effect in the smaller logit specification.

### Big logit model
```{r}
#| echo: false
knitr::kable(summary(m_big_logit)$coefficients, 
             caption = "Big logit model")

```
In the larger logit model, the main pattern remains similar. Gun ownership is still positively associated with voting for Trump, while higher education is negatively associated with the outcome. After adding region, race, type of community, and income, the model captures more variation in the data, suggesting that these background variables are also important for explaining Trump voting.


```{r}
#| echo: false
knitr::kable(summary(m_small_probit)$coefficients, 
             caption = "Small probit model")
```
The small probit model gives results that are very similar to the small logit model. Gun ownership is again positively associated with voting for Trump, while higher education is negatively associated with the outcome. The effect of being female remains negative, but not statistically significant.

```{r}
#| echo: false
knitr::kable(summary(m_big_probit)$coefficients, 
             caption = "Big probit model")
```
The large probit model also shows a pattern similar to the large logit model. The signs and significance levels of the main variables are broadly consistent across the two link functions, which suggests that our conclusions are not very sensitive to whether we use a logit or a probit link.

```{r}
#| echo: false
knitr::kable(anova(m_small_logit, m_big_logit), 
             caption = "ANOVA: Small vs Large logit Model") 
      
```


```{r}
#| echo: false
knitr::kable(anova(m_small_probit, m_big_probit), 
             caption = "ANOVA: Small vs Large probit Model")
```
The ANOVA results show that the larger model improves the fit significantly compared with the smaller model, both for the logit and probit specifications. This suggests that the additional variables contribute useful explanatory information and should be included in the final model.

To compare the overall fit of the four models, we use the Akaike Information Criterion (AIC):
$$
AIC = 2p - 2 \log(L(\hat{\beta}))
$$
where $p$ is the number of parameters in the model and $(L(\hat{\beta})$ is the likelihood evaluated at the estimated parameter values.Lower AIC values indicate a better balance between model fit and model complexity.

```{r}
#| echo: false
aic_tab <- AIC(m_small_logit, m_big_logit, m_small_probit, m_big_probit)
rownames(aic_tab) <- c("Small logit", "Large logit", "Small probit", 
                       "Large probit")

knitr::kable(aic_tab[, c("df", "AIC")], caption = "AIC")


```
Overall, the results suggest that demographic and social background variables are important for explaining Trump voting in 2020. In particular, gun ownership is associated with a higher probability of voting for Trump, while higher education is associated with a lower probability. Both the ANOVA results and the AIC values indicate that the larger models fit the data better than the smaller models. Since the large logit model has the lowest AIC among the four models and is easier to interpret than the corresponding probit model, we choose it as our preferred final model. The similarity between the logit and probit results also suggests that our main conclusions are robust to the choice of link function.

```{r}
#| output: false
#| echo: false
nobs(m_small_logit)
nobs(m_big_logit)
nobs(m_small_probit)
nobs(m_big_probit)
```

# Task 3 
To answer this task, we first created a new dataset where we only kept observations with complete information on the variables used in our models: sex, gun ownership, education, region, race, type of community, and income. We did this by using complete.cases(...), which checks which rows have no missing values (NA) on these variables. This step is necessary because the models cannot make predictions for individuals if some of the explanatory variables are missing.

After this, we selected the individuals in this cleaned dataset who belong to the Midwest and who did not answer the question about the 2020 presidential election. In the dataset, Midwest is coded as region == 2, so we use this condition to identify that region. At the same time, we use is.na(pres20) to find the non-respondents, that is, the individuals with a missing value on the outcome variable. The symbol & is used to combine these conditions, meaning that both must be true at the same time: the person must both live in the Midwest and have a missing value on pres20.

Finally, we used our two models, the larger logit model and the larger probit model, to predict the probability that each individual in this group voted for Trump. We then took the average of these predicted probabilities. This gives us a model-based estimate of the proportion of Trump voters among the non-respondents in the Midwest.


```{r}
#| echo: false
GSS_pred <- GSSdata24[
  complete.cases(GSSdata24[, c("sex", "owngun", "degree", "region", 
                               "race", "res16", "realinc")]),
]

midwest_nr <- GSS_pred[GSS_pred$region == 2 & is.na(GSS_pred$pres20), ]

logit_est <- mean(predict(m_big_logit, newdata = midwest_nr, 
                          type = "response"))
probit_est <- mean(predict(m_big_probit, newdata = midwest_nr, 
                           type = "response"))

pred_tab <- data.frame(
  Model = c("Large logit", "Large probit"),
  Estimated_percent = c(logit_est, probit_est) * 100
)

knitr::kable(pred_tab, digits = 1, caption = "predictions for non-respondents in the Midwest")


```
The large logit model estimates that approximately 31.6 percent of the non-respondents in the Midwest voted for Trump, while the large probit model gives a very similar estimate of 31.8 percent. Since the two estimates are almost identical, the result appears robust to the choice of link function. We therefore conclude that the predicted proportion of Trump voters among non-respondents in the Midwest is about 32 percent.


# Task 4
Saknas:
- Evaluate the censoring of data. 
- Goodness of fit testet behöver nödvändigtvis inte vara överdepresion?
- Kolla igenom språket
- Se till att many_partners matchar modellerna
- Längre text om Zelterman - Se till att den är korrekt?
- Kolla mallformatering så att allt stämmer. 

```{r}
#| echo: false
#| output: false
table(GSSdata24$partners , useNA = 'always')
```
### Data description and censoring of data:
The variable `partners` describes the number of sexual partners reported by different individuals. The outcome variable is partially censored and grouped (may be because of sensitive information), since some categories have intervals rather than exact counts:

- Category 5: Corresponds to 5-10 partners,

- Category 6: Corresponds to 11+ partners,

- Category 9: Corresponds to 1 or more partners, unspecified. 

Generalized Linear Models (GLM) require an observed response variable because likelihood estimation relies on realized outcome. Since standard GLM do not directly fit grouped censoring, an approximation of the count variable was required. We therefore constructed an approximate numerical outcome:

$$Y_i = Numbers\ of\ partners$$
We approximate the groping by:

- Category 5 --> 7 partners (midpoint approximation),

- Category 6 --> 12 partners (lower bound approximation),

- Category 9 --> NA (missing value). 

Observations with missing values were removed. This approximation introducing measurements errors, but preserves ordering and keeps the overall distributional structure. 

We created a binary variable, if married or not: 

$$married_i = \begin{cases} 1 & \text{if married} \\ 0 & \text{otherwise} \end{cases}$$
```{r}
#| echo: false
dat_24 <- GSSdata24
dat_24$partners_num <- dat_24$partners

dat_24$partners_num[dat_24$partners == 5] <- 7
dat_24$partners_num[dat_24$partners == 6] <- 12
dat_24$partners_num[dat_24$partners == 9] <- NA

dat_24 <- subset(dat_24, !is.na(partners_num))

dat_24$married <- dat_24$marital == 1
```

## Model 1 - Poisson regression (log-link):
For our first model we decided to use a poisson regression with a log link function, because the response represents counts of data. 

$$log(\mu_i) = \beta_0 + \beta_1 \cdot married_i$$
Where: 
$$E(Y_i) = \mu_i $$
Therefore the model explains the expected number of partners. 

```{r}
#| echo: false
pois_1 <- glm(partners_num ~ married,
              family = poisson(link = "log"),
              data = dat_24
)

knitr::kable(summary(pois_1)$coefficients, 
             caption = "Poisson Model")

b <- coef(pois_1)

mu_unmarried <- exp(b[1]) # = 1.124277

mu_married <- exp(b[1] + b[2]) # = 1.037569

odds_ratio_b1 <- exp(b[2])

expected_num <- data.frame(Status = c("Unmmaried", "Married"), 
                           expected_num = c(mu_unmarried, mu_married) 
                           )
knitr::kable(expected_num, 
             caption = "Expected Numbers Of Partners")
```

### Interpretation - Poisson:
Intercept (0.11714) represent:
$$log(\mu_{unmarried})$$
After exponentiation gives the expected number of partners among unmarried individuals ($e^{\beta_0} ≈ 1.124$), shown in the table 13. The coefficient is statistically significant at $1\%$ level.

The coefficient `marriedTRUE` (-0.08026) tells us how being married affects the number of partners. / Multiplier if married: $e^{\beta_1} = 0.923$ / A negative coefficient implies that married individuals reports fewer partners on average. The estimated affect gives us a approximately $8\%$ reduction in expected number of partners if being married. The coefficient is statistically significant at $5\%$ level. Shown in table 13, the expected number of partners for someone who is married is $e^{\beta_0 + \beta_1} ≈ 1.038$. 

### Poisson assumption and overdispersion:

The Poisson model assume:
$$ Var(Y_i) = E(Y_i)$$
```{r}
#| echo: false
pois_mean <- mean(dat_24$partners_num)
pois_var <- var(dat_24$partners_num)

assumttion_pois <- data.frame(Measurments = c("Mean", "Variance"), 
                           Value = c(pois_mean, pois_var) 
                           )
knitr::kable(assumttion_pois, 
             caption = "Model assumtions")
```
However, as shown in table 14 the variance (2.78) exceed the mean value (1.09) which indicates overdispersion. This has important consequences for the model, as it likely leads to underestimated standard errors which makes the significance tests look more reliable than they actually are. 

### Evaluation of Poisson model - Godness Of Fit:
To evaluate the fit of the poisson model we conducted a deviance-Goodness of fit test. The model suggests, as above, that the mean is equal to the variance, but as we concluded that is not the scenario. For our test we started by building our hypotheses:

$$H_0 : \text{the Poisson model is adequately specified}$$
$$H_a : \text{the Poisson model is not adequately specified}$$
```{r}
#| echo: false
p_god_test <- pchisq(deviance(pois_1), 
                     df.residual(pois_1),
                     lower = FALSE)

cat("Goodness of fit - p-value:", p_god_test)

```
### Interpretation - Goodness Of Fit test:
The test returned a p-value of $8.967 \cdot 10^{-48}$. Since this value is far below any standard significance level (p > 0.001) we reject the null hypothesis. This result conduct that there is overdispersion in the data, meaning the variation in the number of partners is significantly greater than the Poisson model can predict. Because of this, the Poisson model is best viewed as a "starting model" rather than a perfect fit for the data. The result may suggest a reason to use alternative models instead. By splitting the analysis into two steps, we can better capture the different behaviors in data that a single poisson model might miss out. 

## Model 2 - Binary Models (logit-Link):

The distribution of partners, and evaluation of poisson model, suggests that behavior occur in stages rather than as a single counting process. Instead of a counting process, we focus on the binary stages between having at least one partner, and having more than one partner. We therefor apply a logit-link to model the probability of having `any` or `many` partner/s. We constructed two binary outcomes:

- Having at least one partner (any): $$Y_1 = (Y>0)$$

- Having more than one partner (many): $$Y_2 = (Y>1)$$

We created a Logistic Regression model:
$$ g(p(x))=log(\frac{p(x)}{1 - p(x)}) = \beta_0 + \beta_1 \cdot married_i = \eta_i$$

```{r}
#| echo: false
dat_24$any_partner <- dat_24$partners_num > 0

dat_24$many_partners <- dat_24$partners_num > 1
```

### Logit model - at least one partner:

```{r}
#| echo: false
logit_any <- glm(any_partner ~ married, 
                 family = binomial(link = "logit"), 
                 data = dat_24
                 )
knitr::kable(summary(logit_any)$coefficients,
             caption = "Logit Model - at least one partner")

b_2 <- coef(logit_any)

odds_any_unmarried <- exp(b_2[1]) # 1.46
odds_any_married <- exp(b_2[1] + b_2[2]) # 7.238

prob_any_unmarried <- odds_any_unmarried / (1 + odds_any_unmarried) # 0.5934
prob_any_married <- odds_any_married / (1 + odds_any_married) # 0.8786

prob_any <- data.frame(Status = c("Unmmaried", "Married"), 
                           prob_any = c(prob_any_unmarried, prob_any_married),
                       odds_any = c(odds_any_unmarried, odds_any_married)
                       )
knitr::kable(prob_any, 
             caption = "Probability/Odds - at least one partner")
```
### Interpretation - Logit - at least one partner:

Exponentiated coefficients represent odds ratios:
$$\text{Odds Ratio} = e^{\beta_i}$$

Intercept (0.378): Representing the log-odds for unmarried individuals and corresponds to an odds of 1.46, or a 59 percent probability of having at least one partner. 

The coefficient for `marriedTRUE` is positive (1.6) and highly significant (p-value < 0.001). By adding this coefficient to the intercept and exponentiating $e^{({\beta_0 + \beta_1})}$, we find that married individuals have an odds of 7.24, which corresponds to an 88 percent probability.  

The p-value for both the intercept and the marriage coefficient are small (p < 0.001), indicating that the results are highly statistically significant. Results shows that married individuals have substantially higher probability of having at least one partner. 

### Logit model - more than one partner:

```{r}
#| echo: false
logit_many <- glm(many_partners ~ married, 
                 family = binomial(link = "logit"), 
                 data = dat_24
                 )
knitr::kable(summary(logit_many)$coefficients,
             caption = "Logit Model - more than one partner")

b_3 <- coef(logit_many)

odds_many_unmarried <- exp(b_3[1]) # 0.19076
odds_many_married <- exp(b_3[1] + b_3[2]) # 0.039

prob_many_unmarried <- odds_many_unmarried / (1 + odds_many_unmarried) # 0.16
prob_many_married <- odds_many_married / (1 + odds_many_married) # 0.0376

prob_many <- data.frame(Status = c("Unmmaried", "Married"), 
                           prob_many = c(prob_many_unmarried, prob_many_married),
                       odds_many = c(odds_many_unmarried, odds_many_married)
                       )
knitr::kable(prob_many, 
             caption = "Probability/Odds - more than one partner")
```

### Interpretation - Logit - more than one partner:

Intercept (-1.65673): Representing the log-odds for unmarried individuals and corresponds to an odds of 0.19, or a 16 percent probability of having more than one partner. 

The coefficient for `marriedTRUE` is negative (-1.59) and highly significant (p-value < 0.001). By adding this coefficient to the intercept, we find that married individuals have an odds of 0.039, which corresponds to an 3.76 percent probability.  

The p-value for both the intercept and the marriage coefficient are small (p < 0.001), indicating that the results are highly statistically significant. Results shows that married individuals have substantially lower probability of having more than one partner.

### Conclusion Logit models:

These models show a clear two way pattern. Marriage at first sight substantially increases the chance of having a partner ($59 \%$ to $88 \%$). However, marriage then substantially decreasing the chance of having multiple partners ($16 \%$ to $3.8 \%$). 
 
## Model 3 - Binary model (Probit link):

To make sure the results are reliable, we compare the Logic models with the Probit models (Tables 15, 17 with 19 and 20). While they use different formulas, they are both used to analyse binary data. Probit:
$$\phi^{-1}(p) = \beta_0 + \beta_1 \cdot married = \eta$$
```{r}
#| echo: false
probit_any <- glm(any_partner ~ married, 
                  family = binomial(link = "probit"),
                  data = dat_24
                  )

probit_many <- glm(many_partners ~ married, 
                  family = binomial(link = "probit"),
                  data = dat_24
                  )
knitr::kable(summary(probit_any)$coefficients,
             caption = "Probit model - at least one partner")

knitr::kable(summary(probit_many)$coefficients,
             caption = "Probit model - more than one partner")

prob_any_unmarried_probit <- pnorm(probit_any$coefficients[1])
prob_any_married_probit <- pnorm(probit_any$coefficients[1] + 
                                     probit_any$coefficients[2])

prob_many_unmarried_probit <- pnorm(probit_many$coefficients[1])
prob_many_married_probit <- pnorm(probit_many$coefficients[1] + 
                                     probit_many$coefficients[2])

prob_probit <- data.frame(Status = c("Unmmaried", "Married"), 
                           prob_any_probit = 
                            c(prob_any_unmarried_probit, prob_any_married_probit), 
                          prob_many_probit = 
                            c(prob_many_unmarried_probit, prob_many_married_probit)
)

knitr::kable(prob_probit, 
             caption = "Probability Probit")
```

In the Probit models, the coefficients represent changes in Z-score rather than odds like in the Logit models.  

### Interpretation - Probit - at least one partner:

Intercept (0.2363): Represents the Z-score for unmarried individuals. It corresponds to a probability of approximately $59 \%$, which matches our earlier Logic model (at least one partner). 

The coefficient `marriedTrue` (0.9318) is positive and highly significant (p < 0.001). It shows that being married increases the Z-score by 0.93 standard deviations. If being married, moves the probability from $59 \%$ to $88 \%$, witch again align with the Logit model. 

### Interpretation - Probit - more than one partner:

Intercept (-0.9936): Represents the Z-score for a unmarried individual. It corresponds to a probability of approximately $16 \%$, which matches our earlier Logic model (more than one partner).

The coefficient `marriedTrue` (-0.786) is negative and highly significant (p < 0.001). It shows that being married decreases the Z-score by 0.79 standard deviations. If being married, drops the probability of having multiple partners from $16 \%$ to $3.8 \%$, witch again align with the Logit model. 

### AIC - Assesment of fit Logit/Probit:
To check how well our models work, we use the AIC. AIC measures the balance between a models accuracy and its complexity, which means that its balances the likelihood and penalizing for unnecessary parameters (p). A lower AIC value means the model is more efficient. Since AIC is most meaningful when models are fitted to the same response variable, we will only compare the logit and probit models within each stage of the analysis. 
$$ AIC = 2p-2 log (L(\widehat{\beta}))$$
```{r}
#| echo: false
aic_tab_2 <- AIC(logit_any, logit_many, probit_any, probit_many)

knitr::kable(aic_tab_2, caption = "AIC Table")
```
### Interpretation - AIC:
As shown in the table, the logit and probit models have identical or nearly identical AIC values within each binary outcome. This suggests that the results are not sensitive to the choice between logit and probit. In other words, the main conclusion remain the same regardless of which of these two link-functions is used. 

The AIC values between `any partner` (2652.328) and `many partners` (1561.441) are different and may be due to that the second stage (more than one partner) looks at a smaller, and more specific group of individuals (those whom already have at least one partner). The AIC results also shows that our two step method may be a good choice, and by splitting the analysis into two steps we can better understand:

i) The probability of having at least one partner, and

ii) The probability of having more than one partner.  

### Prediction For a Hypothetical Respondent:
To provide a clear prediction, we analyze a hypothetical individual under two conditions, unmarried and married, using both Logit and Probit link functions. The results, seen in table 23, are transformed into probabilities using the command (type = "response"). 

```{r}
#| echo: false
new_person <- data.frame(married = c(FALSE, TRUE))

pred_table <- data.frame(married = new_person$married, 
# Poisson (antal)
        pois_count = predict(pois_1, 
        newdata = new_person, 
        type = "response"),
# Logit (any)
        logit_any = predict(logit_any,
        newdata = new_person,
        type = "response"), 
# Logit (many)
        logit_many = predict(logit_many,
        newdata = new_person,
        type = "response"), 
# Probit (any)
        probit_any = predict(probit_any,
        newdata = new_person,
        type = "response"), 
# Probit (many)
        probit_many = predict(probit_many,
        newdata = new_person,
        type = "response")
)
knitr::kable(pred_table, 
             caption = "Predicted Outcomes (response)")

```
The Poisson model predicts a slightly higher expected count of partners for unmarried individuals (1.1243) compared to married individuals (1.038). However the two steps binary models reveal a more detail analysis. 

For the first stage of the analysis, the models predict a 0.5934 probability for an unmarried person to have at least one partner, compared to a significantly higher probability of 0.8786 for a married person. In the second stage the probability of having multiple partners drops from 0.16 for an unmarried person to 0.038 for those who are married. 

Overall the results show that marriage is primary related to the probability of having a partner and decreasing the probability of having more than one partner, rather than determining the exact number of partners an individual has. 

### Zelterman:
This can be related to the general idea behind Zeltermans reasoning, where the distinction between zero counts and positive counts may reflect the different underlying processes. In our case, the difference between having no partners, having at least one partner and having more than one partner appears to be important. Our approach is not a full Zelterman model, but it follows a similar intuition by separating different parts of the response rather than stimulating everything in one single count model.  
