---
title: "Generalized linear models - VT26 "
subtitle: "Inlämningsuppgift 1, 2.5 hp"
author: 
- David Müller
- Filip Ström
- Joel Söderman
date: last-modified
format: 
  pdf: default
  html:
    self-contained: true
language: 
  title-block-author-single: " "
crossref-fig-title: "Figur"
title-block-banner-color: Primary
title-block-published: "Publicerad"
author-title: "Gruppmedlemmar"
title-block-banner: true
editor: source
---

```{r}
#| output: false
#| echo: false
library('foreign')
library('haven') 
library('labelled')

load(url("https://raw.githubusercontent.com/johankoskinen/ST304G/main/data/GSS2022selection.RData"))

load(url("https://raw.githubusercontent.com/johankoskinen/ST304G/main/data/GSS2024selection.RData"))
```

```{r}
#| output: false
#| echo: false
table(GSSdata24$pres20, useNA='always')
table(GSSdata24$pres20==2, useNA='always')


```
# Introduction
This report analyzes data from the General Social Survey (GSS) using generalized linear models. In Part A, we study which factors are associated with voting for Donald Trump in the 2020 presidential election, using data from the 2022 and 2024 surveys. In Part B, we examine the reported number of sexual partners and discuss suitable models for this type of outcome.

Throughout the report, we compare different model specifications and link functions, interpret the estimated effects, and evaluate how well the models fit the data. We also discuss missing values and other limitations where relevant.

### Note on AI use 
We used AI in a limited way to help format the output in the report, mainly by suggesting how to use knitr::kable() and similar commands to present summary() and anova() results more clearly. 

# Task 1
::: {.columns}
::: {.column width="50%"}
```{r}
#| echo: false
#| 
model_24 <- glm( pres20==2 ~ 
                   I(sex==2) + 
                   I(owngun==1) +
                   I(degree %in% c(4,5)), 
                 data = GSSdata24,
                 family = "binomial")


knitr::kable(summary(model_24)$coefficients, digits = 3)

```
For the 2024 model, gun ownership is positively associated with voting for Trump, while having a Bachelor’s or Graduate degree is negatively associated with the outcome. The coefficient for being female is negative, but it is not statistically significant, which means that we do not find clear evidence of a gender effect in this model.


```{r}
#| echo: false
model_22 <- glm((pres20 == 2) ~ 
                  I(sex == 2) + 
                  I(owngun == 1) + 
                  I(degree %in% c(4,5)),
               family = "binomial",
               data = GSSdata22)

knitr::kable(summary(model_22)$coefficients, digits = 3)
```
The 2022 model shows a similar pattern. Gun ownership is again positively associated with voting for Trump, while higher education is negatively associated with the outcome. In this model, being female also has a negative coefficient, and here the effect is statistically significant.


To formally test whether the association between being female and voting for Trump changed over time, we then created a pooled dataset that combines the observations from 2022 and 2024. In the code, we first selected the variables needed for the analysis, then added a new variable called year to identify which survey year each observation comes from, and finally stacked the two datasets together using rbind(). In the combined model, we included an interaction term between I(sex == 2) and I(year == 2024). This interaction term is the key part of the code, because it tests whether the effect of being female differs between 2022 and 2024, rather than simply estimating one common effect across both years.
```{r}
#| echo: false
vars_needed <- c("pres20", "sex", "owngun", "degree")

GSSdata22_sub <- GSSdata22[, vars_needed]
GSSdata24_sub <- GSSdata24[, vars_needed]

GSSdata22_sub$year <- 2022
GSSdata24_sub$year <- 2024

GSSdata <- rbind(GSSdata22_sub, GSSdata24_sub)

model <- glm(
  I(pres20 == 2) ~ .,
  data = GSSdata,
  family = binomial
)

knitr::kable(summary(model)$coefficients, digits = 3)
```
The p-value for the interaction term is 0.787. Since this p-value is large, there is no statistical evidence that the association between being female and voting for Trump changed between the two years. Therefore, we cannot reject the null hypothesis that the effect of being female is the same in 2022 and 2024. 

# Task 2:
We estimated four models, a small and a large specification, each with both a logit and a probit link. In the code, we first defined a set of variables that are used in at least one of the models. We then created GSS_clean by applying na.omit() to these variables, which means that all four models are estimated on the same complete cases. This is important, because it makes the comparison between models more fair. If the models were estimated on different subsets of the data, differences in AIC or model fit could partly reflect differences in the sample rather than differences in the models themselves.

The smaller models include only sex, gun ownership, and education, while the larger models also include region, race, type of community, and logged income. We estimated both logit and probit versions in order to compare whether the results depend on the choice of link function. We then used summary() to inspect the coefficients, anova() to compare the smaller and larger models formally, and AIC() to compare overall model fit. Finally, we used nobs() to verify that all four models were estimated on the same number of observations.

```{r}
#| output: false
#| echo: false
vars_needed <- c(
  "pres20", "sex", "owngun", "degree",
  "region", "race", "res16", "realinc"
)

# Behåll bara kompletta observationer
GSS_clean <- na.omit(GSSdata24[, vars_needed])

# Small logit
m_small_logit <- glm(
  pres20 == 2 ~
    I(sex == 2) +
    I(owngun == 1) +
    I(degree %in% c(4, 5)),
  data = GSS_clean,
  family = binomial(link = "logit")
)

# Big logit
m_big_logit <- glm(
  pres20 == 2 ~
    .,
  data = GSS_clean,
  family = binomial(link = "logit")
)

# Small probit
m_small_probit <- glm(
  pres20 == 2 ~
    I(sex == 2) +
    I(owngun == 1) +
    I(degree %in% c(4, 5)),
  data = GSS_clean,
  family = binomial(link = "probit")
)

# Big probit
m_big_probit <- glm(
  pres20 == 2 ~
    .,
  data = GSS_clean,
  family = binomial(link = "probit")
)

```

### Small logit model

```{r}
#| echo: false
knitr::kable(summary(m_small_logit)$coefficients, digits = 3)
```
The small logit model shows that gun ownership is positively associated with voting for Trump, while having a Bachelor’s or Graduate degree is negatively associated with the outcome. The coefficient for being female is negative, but it is not statistically significant, which means that we do not find clear evidence of a gender effect in the smaller logit specification.

### Big logit model
```{r}
#| echo: false
knitr::kable(summary(m_big_logit)$coefficients, digits = 3)

```
In the larger logit model, the main pattern remains similar. Gun ownership is still positively associated with voting for Trump, while higher education is negatively associated with the outcome. After adding region, race, type of community, and income, the model captures more variation in the data, suggesting that these background variables are also important for explaining Trump voting.


### Small probit model
```{r}
#| echo: false
knitr::kable(summary(m_small_probit)$coefficients, digits = 3)
```
The small probit model gives results that are very similar to the small logit model. Gun ownership is again positively associated with voting for Trump, while higher education is negatively associated with the outcome. The effect of being female remains negative, but not statistically significant.

### Big probit model
```{r}
#| echo: false
knitr::kable(summary(m_big_probit)$coefficients, digits = 3)
```
The large probit model also shows a pattern similar to the large logit model. The signs and significance levels of the main variables are broadly consistent across the two link functions, which suggests that our conclusions are not very sensitive to whether we use a logit or a probit link.


### ANOVA: Small vs Large Logit Model
::: {.columns}
```{r}
#| echo: false
knitr::kable(as.data.frame(anova(m_small_logit, m_big_logit, test = "Chisq")), 
             digits = 3)
```

### ANOVA: Small vs Large probit Model
```{r}
#| echo: false
knitr::kable(as.data.frame(anova(m_small_probit, m_big_probit, 
                                 test = "Chisq")), digits = 3)
```
The ANOVA results show that the larger model improves the fit significantly compared with the smaller model, both for the logit and probit specifications. This suggests that the additional variables contribute useful explanatory information and should be included in the final model.

### AIC comparison of the four models 
```{r}
#| echo: false
aic_tab <- AIC(m_small_logit, m_big_logit, m_small_probit, m_big_probit)
rownames(aic_tab) <- c("Small logit", "Large logit", "Small probit", 
                       "Large probit")

knitr::kable(aic_tab[, c("df", "AIC")], digits = 3)


```
Overall, the results suggest that demographic and social background variables are important for explaining Trump voting in 2020. In particular, gun ownership is associated with a higher probability of voting for Trump, while higher education is associated with a lower probability. Both the ANOVA comparisons and the AIC values show that the larger models fit the data better than the smaller models. Since the large logit model has the lowest AIC and is easier to interpret than the corresponding probit model, we choose it as our preferred final model. The similarity between the logit and probit results also suggests that the main conclusions are robust to the choice of link function.

```{r}
#| output: false
#| echo: false
nobs(m_small_logit)
nobs(m_big_logit)
nobs(m_small_probit)
nobs(m_big_probit)
```


# Task 3

Anteckning, tas bort sen: MIDWEST = 2

```{r}
newdata <- data.frame(
  sex = 2, 
  owngun = 1,
  degree = 4,
  region = factor(2),
  race = factor(1),
  res16 = factor(2),
  realinc = 45000)
```


# Task 3 - Joel
To answer this task, we first created a new dataset where we only kept observations with complete information on the variables used in our models: sex, gun ownership, education, region, race, type of community, and income. We did this by using complete.cases(...), which checks which rows have no missing values (NA) on these variables. This step is necessary because the models cannot make predictions for individuals if some of the explanatory variables are missing.

After this, we selected the individuals in this cleaned dataset who belong to the Midwest and who did not answer the question about the 2020 presidential election. In the dataset, Midwest is coded as region == 2, so we use this condition to identify that region. At the same time, we use is.na(pres20) to find the non-respondents, that is, the individuals with a missing value on the outcome variable. The symbol & is used to combine these conditions, meaning that both must be true at the same time: the person must both live in the Midwest and have a missing value on pres20.

Finally, we used our two models, the larger logit model and the larger probit model, to predict the probability that each individual in this group voted for Trump. We then took the average of these predicted probabilities. This gives us a model-based estimate of the proportion of Trump voters among the non-respondents in the Midwest.


```{r}
#| echo: false
GSS_pred <- GSSdata24[
  complete.cases(GSSdata24[, c("sex", "owngun", "degree", "region", 
                               "race", "res16", "realinc")]),
]

midwest_nr <- GSS_pred[GSS_pred$region == 2 & is.na(GSS_pred$pres20), ]

logit_est <- mean(predict(m_big_logit, newdata = midwest_nr, 
                          type = "response"))
probit_est <- mean(predict(m_big_probit, newdata = midwest_nr, 
                           type = "response"))

pred_tab <- data.frame(
  Model = c("Large logit", "Large probit"),
  Estimated_rpercent = c(logit_est, probit_est) * 100
)

knitr::kable(pred_tab, digits = 1)


```
The large logit model estimates that approximately 31.6 percent of the non-respondents in the Midwest voted for Trump, while the large probit model gives a very similar estimate of 31.8 percent. Since the two estimates are almost identical, the result appears robust to the choice of link function. We therefore conclude that the predicted proportion of Trump voters among non-respondents in the Midwest is about 32 percent.



Part B Kommentarer
- Saknar odds ratio
- Deviation korrekt?
- Mer latex som stämmer överrens med frl anteckningarna. 
- Assesment of fit
- Skriv om Zelterman
- Se kommentarer i text. 
- Skriv om till engelska.
- Använda probit link?
- AIC värden
- Figurer när vi predicierar
- Skriva att poission inte är bättre än logit (mean = var)
- Goodness of fit test

# Part B
### Task 4
```{r}
table(GSSdata24$partners , useNA = 'always')
```
Variabeln partners beskriver antal partners som den tillfrågade har rapporterat. Variabeln är delvis censorerad då vissa kategorier representerar ett intervall (ex 5 som motsvarar 5-10 i antal) snarare än ett exakt antal. För att hantera detta problem behövde vi approximera ett ungefärligt exakt värde för dessa intervall. Vi ansatte kategori 5 till 7 stycken (i antal) då detta låg nära ett "medel" i kategorin samt 12 för kategorin 11+ (ligger nära 11). Kategori 9 innebär "ospecificierat" och vi valde att sätta denna till NA (saknat värden för att inte snedvrida resultatet). 

### Mittpunktsapproximation (begrepp vi använder?)

Sedermera valde vi att ta bort de saknade värderna från vårt dataset då GLM modeller kräver observerade responsvärden. Dessutom så baseras GLM estimeringar på likelihoodfunktioner vilka kräver observerade utfall. GLM-modeller kan heller inte modellera censorerad data vilket således gör att vi måste approximera intervallen för att möjliggöra likelihood estimationer. 

### Kan tas bort
Mätfel:
I och med att vi approximerar de grupperade variablerna så tillkommer det mätfel, dessa kan vi inte udgås men för att möjliggöra GLM krävs dessa antaganden. Detta då vi dels vidhåller samma ordning i datan samt att vi förhåller oss till "top coded" problematiken. Hämtat från Wikipedia (2020-12-02, "Top-coded"): "To help overcome this problem, CPS provides the mean value of top coded values". 

Vi skapar även en binär variabel "married" för att kunna undersöka om äktenskap påverkar antal partners: TRUE = gift, FALSE = ej gift. 

```{r}
dat <- GSSdata24
dat$partners_num <- dat$partners

dat$partners_num[dat$partners == 5] <- 7
dat$partners_num[dat$partners == 6] <- 12
dat$partners_num[dat$partners == 9] <- NA

# Kanske lite längre förklaring till denna kod-del. 
dat <- subset(dat, !is.na(partners_num))

dat$married <- dat$marital == 1
```

### Länkfunktion 1 - Poission Log-link:
Eftersom att responsvariabeln representerar antal har vi valt att använda en poisson regression med log länk. 

$$log(\mu) = \beta_0 + \beta_1 * marriedTRUE$$
där:
$$E(Y) = \mu $$
Alltså, vi modellerar förväntat antal partners. 
```{r}
pois_1 <- glm(partners_num ~ married,
              family = poisson(link = "log"),
              data = dat
)

summary(pois_1)

b <- coef(pois_1)

# Förväntat antal partners för en ogift person
mu_unmarried <- exp(b[1]) # = 1.124277

# Förväntat antal partners vid gift
mu_married <- exp(b[1] + b[2]) # = 1.037569

list("partners vid ogift" = mu_unmarried, "partners vid gift" = mu_married)
```

Tolkning av koefficienter:

Intercept (b[1] = 0.11714): Detta är det logaritmerade "mu" för en ogift individ När vi transformerar detta (exp) får vi ut att det förväntade antalet partners för en ogift individ uppgick till 1.124277 stycken. 

marriedTRUE (-0.08026): Detta anger skillnaden mellan att vara ogift kontra gift. Vid transformation av detta värdet tillsammans med interceptet får vi ut ett förväntat värde på 1.0376 Sätt ur en procentuell vinkel så kan vi säga att gifta personer i genomsnitt har $8 \%$ färre förväntade partners än ogifta. (Se nedan varför vi kan göra denna tolkning)

p-värde (0.0426): marriedTRUE är signifikant på 95% nivån, således kan vi med 95% säkerhet konstatera att gifta individer har färre partners. 

Modellkontroll:
```{r}
# Korrekt?
deviance(pois_1) / df.residual(pois_1) # 1.474442
```
Här testar vi huruvida det finns någon overdipersion. Vi antar i en poisson fördelning att variansen är lika med medelvärdet. Kvoten mellan deviance och frihetsgraderna indikerar på att där finns overdipersion, vilket kan innebära att poissonmodellen kan underskatta variationen i data. 

Overdipersion är rimligt i detta fall då vi har aproximerat samt hanterat censurerad data. Overdipersion > 1

### AIC eller goodness of fit?
### Vi saknar en modellutvärdering

### Länkfunktion 2 - Logit-modell:

$$Odds = e^{\beta}$$
$$Sannolikhet = \frac{Odds}{1 + Odds}$$
Vi skapar 2 stycken binära variabler:

- Minst en partner (any_partner)

- Fler än en partner (many_partners)
```{r}
dat$any_partner <- dat$partners_num > 0

dat$many_partners <- dat$partners_num > 1
```

```{r}
# Logit modell - minst en partner:
logit_any <- glm(any_partner ~ married, 
                 family = binomial(link = "logit"), 
                 data = dat
                 )
summary(logit_any)

b_2 <- coef(logit_any)

odds_any_unmarried <- exp(b_2[1]) # 1.46
odds_any_married <- exp(b_2[1] + b_2[2]) # 7.238

prob_any_unmarried <- odds_any_unmarried / (1 + odds_any_unmarried) # 0.5934
prob_any_married <- odds_any_married / (1 + odds_any_married) # 0.8786

list("Prob if unmarried" = prob_any_unmarried, 
     "Prob if married" = prob_any_married)
```
Tolkning av koefficienter:

Intercept (0.378): Är vårt logodds för ogifta individer som har haft minst en partner. Vid transformation av detta får vi ut oddset som uppgick till 1.46. Sedermera omvandlar vi detta till en sannolikhet som säger oss att ogifta personer har ca $59 \%$ sannolikhet att ha haft minst en partner

marriedTRUE: Är förändringen i logodds mellan individer som är ogifta och gifta. Vi transformerar detta till odds genom att addera $\beta_1$ (1.6) med interceptet för att sedan exponentiera det (Odds = 7.238). Sedan omvandlar vi oddset till en sannolikhet (0.8786). Här kan det konstateras att gifta individer har ca $88\%$ sannolikhet att ha haft minst en partner, vilket således är mycket högre än de ogifta individerna. Notering: I och med att koeficienten är positiv vet vi att "vara gift" ökar sannolikheten att minst haft en partner, värt att nämna är att detta resultat är ytterst signifikant (p-värde: 2e-16). 

```{r}
# Logit modell - fler än en partner:
logit_many <- glm(many_partners ~ married, 
                 family = binomial(link = "logit"), 
                 data = dat
                 )
summary(logit_many)

b_3 <- coef(logit_many)

odds_many_unmarried <- exp(b_3[1]) # 0.19076
odds_many_married <- exp(b_3[1] + b_3[2]) # 0.039

prob_many_unmarried <- odds_many_unmarried / (1 + odds_many_unmarried) # 0.16
prob_many_married <- odds_many_married / (1 + odds_many_married) # 0.0376

list("Prob if unmarried" = prob_many_unmarried, 
     "Prob if married" = prob_many_married)
```
Tolkning av koefficienter:

Intercept (-1.65673): Är vårt logodds för ogifta individer som har haft fler än en partner. Vid transformation av detta får vi ut oddset som uppgick till 0.19076. Sedermera omvandlar vi detta till en sannolikhet som säger oss att ogifta personer har ca $16 \%$ sannolikhet att ha haft fler än en partner.

marriedTRUE: Är förändringen i logodds mellan individer som är ogifta och gifta. Vi transformerar detta till odds (0.039) och sedan omvandlar detta till en sannolikhet (0.0376). Här kan det konstateras att gifta individer har ca $3.76\%$ sannolikhet att ha haft fler än en partner, vilket således är mycket lägre än de ogifta individerna. Notering: I och med att koeficienten är negativ vet vi att "vara gift" minskar sannolikheten att ha haft fler än en partner, värt att nämna är att detta resultat är ytterst signifikant (p-värde: 2e-16). 

### Kan tas bort (kanske skrivs längre ner.)
Tolkning Länkfunktion 2:
Efter modelleringen kan vi konstatera att äktenskap (att vara gift) ökar sannolikheten att ha haft minst en partner, men att ingå i ett äktenskap samtidigt minskar sannolikheten att ha flera partners. 

### Prediktion (Hypotetisk respondent):
Vi betraktar en hypotetisk respondent som antingen är gift eller ogift och undersöker hur modellerna predicierar antal partners (poisson) respektive sannolikheten (logit). Detta gör vi genom att skapa två hypotetiska individer (new_person) som vi sedan applicerar på våra modeller i en dataframe. Detta skapar en tabell som ger oss de önskade värderna. 

Kommandot type = "response" transformerar modellens linjära prediktor tillbaka till dess originalskala. För poissonmodellen innebär det "förväntat antal partners", medan för logitmodellen innebär det sannolikhet. 

```{r}
new_person <- data.frame(married = c(FALSE, TRUE))

data.frame(married = new_person$married, 
# Poisson (antal)
        antal_partners = predict(pois_1, 
        newdata = new_person, 
        type = "response"),
# Logit (Sannolikhet) > 0
        prob_any_partner = predict(logit_any,
        newdata = new_person,
        type = "response"), 
# Logit (sannolikhet) > 1
        prob_many_partners = predict(logit_many,
        newdata = new_person,
        type = "response")
)

```
Enligt poisson modellen förväntas en ogift individ (1.1243) rapportera något fler partners kontra en gift individ (1.0376). Logit modellen predicerar att gifta individer (0.8786) har högre sannolikhet att ha haft minst en partner jämfört med ogifta individer (0.5934). Därtill kan det konstateras att gifta individer (0.0376) har lägre sannolikhet att ha haft fler än en partner kontra en ogift individ (0.16). Data framen illustrerar tre olika aspekter utifrån den binära variablen married. 

```{r}
#| fig-width: 5
#| fig-height: 3
#| out-width: "50%"
#| fig-align: center
#| echo: false

barplot(table(dat$partners_num), 
     main = "Fördelning över # partners_num", 
     xlab = "Antal partners", 
     ylab = "frekvens", 
     col = "lightblue"
     )
```

Fördelningen av antal partners är högerskev, där många tillfrågade rapporterat låga värden och få höga värden. Detta innebär att att sambandet mellan äktenskap och antalet partners inte är homogent över hela fördelningen. Denna analys blir däremot lite tvetydig då viss del av datan är censurerad. Vi antar däremot en högerskev fördelning. 

Med detta antagande delar vi upp analysen i två delar: 
1) Sannolikheten att ha haft minst en partner, och

2) givet minst en partner, vad är sannolikheten att ha fler än en partner. 

Skillnaden mellan att ha minst en partner och fler än en visar således att äktenskap främst är relaterad till sannolikheten att överhuvudtaget ha en partner, snarare än till ett högre antal partners bland individer som redan haft en partner. I och med detta kan en uppdelad modellstruktur motiveras. 

Detta kan kopplas till Zeltermans modell, som fokuserar på:







